---
title: "GxE Analysis Workshop"
author: "Ellena Girma"
date: "2022/05/27 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
      slideNumberFormat: "%current%"
      ratio: 16:9
---
class: middle, center

```{r setup, include=FALSE }
knitr::opts_chunk$set(echo = TRUE, fig.retina=3, fig.height = 7)

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r style-share-again, echo=FALSE}
xaringanExtra::use_share_again()

```


```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard(
  button_text = "Copy to clipboard",
  success_text = "Copied",
  error_text = "Press Ctrl + C para to copy"
)
```

### Hello, Everyone! 

--

#### My name is Ellena Girma and I am a breeding data systems and analytics specialist at the Alliance Bioversity International - CIAT.

--

#### The goal of this presentation is to demonstrate how R can be used to run GxE analyses.

--

#### I will take you through the workflow that I use to run my GxE analysis for multi environment trials. For each part, I will cover the main functions that I use and what outputs/results you can expect to get.

---

## Introduction to GxE Studies 

.pull-left[
#### Understanding Genotype-Environment (GxE) Interaction
* GXE interaction: the interaction of genotypes and environmental conditions and its joint effect on the traits of interest

* Is there a pattern of phenotypic expression of a trait across various environmental conditions? 
]

--

.pull-right[
#### Selecting & recommending genotypes that have wider or more specific adaptation across various environmental conditions

* Stability: Does the performance of the genotype stay the same across environments?

* Yield Production: Does the genotype perform well across environments? 
]

---

## My Workflow

#### In this presentation I will be sharing the process I use to get from raw data to results. I will be focusing on the response variable yield, but the code I use can be altered for other response variables too. 

Workflow:

* Data Preparation 

* Exploratory Data Analysis

* Means and High Yielding Genotypes & Locations

* Analysis of Variance (ANOVA)

* GGE Biplots

* Stability Analysis

* Additive Main-Effects Multiplicative Interaction (AMMI)

---
class: middle, center

## Let's take a quick tour of R. 

---
class: middle, center

## For this workshop, we'll be focusing on the metan package.

---

## Loading Packages and Importing Data

```{r set-up-workspace-hidden, include = FALSE}
library(xaringan)
library(xaringanthemer)
library(xaringanExtra)
library(here)

```

```{r importing-and-loading, message = FALSE, warning=FALSE}

#loading metan package into workspace
#for analysis
library(metan)

#loading tidyverse package into workspace
#for data manipulation functions
library(tidyverse)

#importing a csv file using the read.csv() function
bean_data <- read.csv(here("GxE-June2022", "csvfile.csv"), 
                      header=TRUE,
                      colClasses = c(rep("factor", 3), rep("numeric", 5)))

```

---
class: middle

## Exploratory Data Analysis (EDA)

### Exploring and summarizing datasets using summary statistics and plots/data visualization methods.

--

Some EDA tasks:

* finding outliers and missing values

* looking at how our data is distributed

* looking at the relationship between variables

* getting summary statistics for each variable of interest

---

## Summary Statistics

.panelset[
.panel[.panel-name[R Code]
```{r summary_stats-p1}
#descriptive stats using metan package
ds_all <- desc_stat(bean_data, stats= "main") 

```
]

.panel[.panel-name[desc_stats output]
```{r summary_stats-pt3}

print(ds_all)

```
]
]

---

#### Checking for outliers: Plots

.panelset[
.panel[.panel-name[Code]
This is for the variable yield. To get potential outliers for another variable, set the var argument to the variable name.
```{r yield-outliers-plotcode, fig.show='hide'}
#Checking for potential outliers in yield variable
find_outliers(bean_data, var = yield, plots =T)
```
]
.panel[.panel-name[Plot]
```{r yield-outliers-plots}
find_outliers(bean_data, var = yield, plots =T)
```
]
]

---

### Checking for Outliers: Looking at Data Points

Let's take a look at which data point are the outliers. Are they all the same location or same genotypes? 

.panelset[
```{r yield-outliers, panelset = TRUE}
#potential outliers
yield_outliers <- c(10, 28, 32, 35, 58, 66, 68, 69, 74 ,75, 84, 93, 109, 127, 131, 134, 157, 165, 167, 168, 173, 174, 183, 192, 208, 209, 226, 230, 233, 256, 264, 266, 267, 272, 273, 282, 291)

yo <- bean_data %>%
  slice(yield_outliers) %>%
  arrange(Genotype) %>%
  slice(1:15)

yo
```
]

---

## Grand Mean Yield of Genotypes Across Locations

We can see the grand mean yield for each genotype across all three locations using the means_by() function.  

.panelset[
.panel[.panel-name[R Code]
```{r grand-yield-means-p1}
#getting means by Genotype
mg <- means_by(bean_data, Genotype)

#Creating a new object with Ordered Means by Genotype
mg_y <- mg %>%
  select(Genotype, yield) %>%
  arrange(desc(yield))
```
]

.panel[.panel-name[Table]
```{r grand-yield-means-p2}
slice(mg_y, 1:10)
```
]
]

---

## Yield Means Across Genotypes at Each Location

Now, let's take a look at the grand mean at each location using the means_by() function. 

.panelset[
.panel[.panel-name[Code]
```{r yield-means-location}
#getting means for all variables by Location
me <- means_by(bean_data, Site)

#Creating a new object with Ordered Means by Location
me_y <- me %>%
  select(Site, yield) %>%
  arrange(desc(yield))
```
]

.panel[.panel-name[Output]
```{r yield-means-location-print}
me_y
```
]
]

---

## Top Performers at Each Location

We can use the ge_winners() function to get the top performing genotype at each location across all response variables. 

```{r top-genotype-across-loc}
#top performing genotype in each env for yield
win <- ge_winners(.data = bean_data, 
                  env = Site, 
                  gen = Genotype, 
                  resp = c(DF, pods.plant, seeds.pod, sw, yield), type = "winners")
```

```{r print-top-genotype-across-loc}
print(win)
```

---
class: middle

## Analysis of Variance (ANOVA): Introduction 

#### Running an ANOVA allows us to 

* Evaluate differences in means among groups

    * see if genotype main effect, environment main effect and GxE interaction effect are significant

    * see which effect contributes most to the variation in performance of the different traits
* Divide the variance into parts to estimate heritability

#### The metan package provides <a href = "https://tiagoolivoto.github.io/metan/reference/index.html">many functions</a> that can be used to run ANOVA. Different functions are available based on the ANOVA type (one-way, two-way), type of model (fixed or mixed effect), etc. 

---

## ANOVA for RCBD

#### anova_joint() is the function used for combined site analysis.

#### get the anova table for the response variable you want through anova_object_name$var_name$anova to 

```{r hidden-assignment, include=FALSE}
rcbd_anova <- anova_joint(bean_data, env = "Site", gen = "Genotype", rep = "Rep", resp = everything())
```

.panelset[
.panel[.panel-name[R Code]
```{r rcbd-anova, results='hide', warning=FALSE}

#Running joint anova for rcbd layout/plot design for all repsonse variables

anova_joint(bean_data, env = "Site", gen = "Genotype", rep = "Rep", resp = everything())
```
]
.panel[.panel-name[Yield ANOVA Table]
```{r Yield-ANOVA-Table}
#getting the ANOVA table for yield 
rcbd_anova_yield <- rcbd_anova$yield$anova
rcbd_anova_yield
```
]
.panel[.panel-name[SW ANOVA Table]
```{r SW-ANOVA-Table}
#getting the ANOVA table for 100g seed weight
rcbd_anova$sw$anova

```
]
.panel[.panel-name[DF ANOVA Table]
```{r DF-ANOVA-Table}
#getting the ANOVA table for days to 75% flowering
rcbd_anova$DF$anova

```
]
]

---

## Estimating Heritability

#### Variance components can be calculated using the results from the ANOVA table. Using these variance components, we can estimate heritability.
#### Use indexing to get the correct value from the anova table: anova_table[row_number, col_number]

.panelset[
.panel[.panel-name[Calc Values 1]

V of line = (Mean sq of treat - Mean sq of site:treat) / (site*rep)

V of line*site = (Mean sq of site:treat - Mean sq of residual) / (rep)

V of error = mean sq of residuals

H2= Vline / Vline + Vline*site + Verror

H2L= Vline/ (line + (Vline*site/nsites) + (Verror/nsite*nrep) 

]
.panel[.panel-name[Calc Values 2]
```{r variance-components-p1}
#variance of line
Vl.yd <- (rcbd_anova_yield[3,4] - rcbd_anova_yield[4,4])/(3*3)

#variance of line by site
Vls.yd <- (rcbd_anova_yield[4,4] - rcbd_anova_yield[5,4])/(3)

#variance of error
Ve.yd <- rcbd_anova_yield[5,4]

#Broad sense hertitability 
H2.yd <- Vl.yd/(Vl.yd + Vls.yd + Ve.yd)

#Hertitability of Line Means
H2L.yd <- Vl.yd/(Vl.yd + Vls.yd/3 + Ve.yd/(3*3))

```
]

.panel[.panel-name[Combing Values into Table]
```{r variance-components-p2}
comb.site.vc <- cbind(rbind(Vl.yd, Vls.yd, Ve.yd, H2.yd, H2L.yd))

colnames(comb.site.vc) <- c("Yield")
rownames(comb.site.vc) = c("V of Line", "V of Line x Site", "V of Error", "Broad Sense Heritability", "Heritability of Line Means")


```
]

.panel[.panel-name[Final Output]
```{r variance-components-table}
print(comb.site.vc)
```
]
]

---
class: middle

## Genotype plus genotype-by-environment (GGE) Models

--

#### Use the gge() function to create a gge model. 

--

#### First, the centering argument needs to be set to "environment" for us to get a GGE model. 

--

#### Then the svp argument can be used to create three types of gge models. Below are the three options: 

* svp = "environment"

* svp = "genotype"

* svp = "symmetrical"


---
class: middle, center

## Genotype plus genotype-by-environment (GGE) Biplots

--

#### Use the plot() function to create plots from your gge models. 


#### Use the type argument to choose the kind of plot you want. There are 10 types to choose from. 



---

## Type argument in plot

.pull-left[

1 = Basic biplot.

2 =  Mean performance vs. stability (gge biplots) or the The Average Tester Coordination view for genotype-trait and genotype-yield*trait biplots.

3 = Which-won-where.

4 = Discriminativeness vs. representativeness.

5 = Examine an environment (or trait/yield*trait combination).
]
.pull-right[
6 = Ranking environments (or trait/yield*trait combination).

7 = Examine a genotype.

8 = Ranking genotypes.

9 = Compare two genotypes.

10 = Relationship among environments (or trait/yield*trait combination).
]

#### Take a look at the <a href = "https://tiagoolivoto.github.io/metan/reference/plot.gge.html">plot documentation</a> page for more information. 

---
class: middle, center

# GGE Model 1
#Let's start with svp = "environment"

---

```{r creating-gen_id_codes, include=FALSE}
# Creating new factor 

bean_data$Gen_id <- as.numeric(factor(bean_data$Genotype, levels =  unique((bean_data$Genotype))))

gen_id_codes <- bean_data %>%
  distinct(Genotype, Gen_id)
```

.panelset[
.panel[.panel-name[Create Model]
```{r create-gge-model}
gge_model <- gge(.data = bean_data,
                 env = Site,
                 gen = Gen_id,
                 resp = yield,
                 centering = "environment",
                 svp = "environment")

```
]
.panel[.panel-name[Create Plot]
```{r basic-biplot-hidden, warning = FALSE, message=FALSE, fig.height = 15, include=FALSE}
#creating the plot
gge_biplot <- plot(gge_model, #<<
                   
                   plot_theme = theme_xaringan(css_file = here("GxE-June2022/xaringan-themer.css")),
  text_font_size = 12,
  title_font_size = 14)+
  ylim(-4500,3000)

```

Basic Biplot is a PC2 vs PC1 plot. Each environment will have a vector starting from the origin. The angle between environments gives you information about the correlation between the environments. 

```{r basic-biplot, results='hide', warning=FALSE, eval=FALSE}
#creating the basic biplot
plot(gge_model, type = 1)
```
]
.panel[.panel-name[Basic Biplot]
```{r plot-basic-biplot, message=FALSE, warning=FALSE}
#call the name of the plot to view it
gge_biplot

```
]
]

---

.panelset[
.panel[.panel-name[R Code]
The discriminativeness vs. representativeness plot is created by setting the type argument to 4.


```{r creating-dvr-hiden, warning=FALSE, message=FALSE, include=FALSE}
dvr <- plot(gge_model, type = 4, #<<
            
             plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14)) +
  ylim(-4500,3000)
```
```{r creating-dvr, results='hide', eval=FALSE}
plot(gge_model, type = 4)
```
]
.panel[.panel-name[Plot]

Selian has the longest vector. We can look at yield performance at Selian and get more information about the yield of the genotypes than at SUA or Uyole. Uyole has the smaller angle to the Average Environment Axis (AEA). This makes it the most representative environment. 

```{r plot-dvr, message=FALSE, warning=FALSE}
dvr
```
]
]

---

.panelset[
.panel[.panel-name[R Code]

Let's compare the environments using another biplot. 
Set the type to 6 to rank environments and set type to 10 to visualize a clear relationship of the environments. 

```{r comparing-biplots, message=FALSE, results='hide', eval=FALSE}
#creating plot 1: ranking environments
plot(gge_model, type=6)

#creating plot 2:relationship among environment
plot(gge_model, type = 10)

```

```{r comparing-biplots-hidden, message=FALSE, include=FALSE}
#creating plot 1: ranking environments
ranking_envrionment <- plot(gge_model, type=6, 
                            plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14)) +
  ylim(-4000,2000)

#creating plot 2:relationship among environment
relationship <- plot(gge_model, type = 10, #<<
                     
                     plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14)) +
  ylim(-4000,2000)

```
]
.panel[.panel-name[Ranking Environment]

```{r ranking-plot, message=FALSE, warning=FALSE}
ranking_envrionment
```
]
.panel[.panel-name[Relationship Between Environments]
```{r comparing-env-biplots, message=FALSE, warning=FALSE}
relationship
```
]
]

---
class: middle, center

# Now, for the second model, let's try svp = "genotype"

---

```{r create-gge-model-2}
gge_model_gen <- gge(.data = bean_data,
                 env = Site,
                 gen = Gen_id,
                 resp = yield,
                 centering = "environment",
                 svp = "genotype")

```

---

##### Svp = "genotype" Biplots

.panelset[
.panel[.panel-name[R Code]
explain plot types

```{r genotype-biplots-code, warning=FALSE, message=FALSE, eval=FALSE}

# mean performance vs stability 
plot(gge_model_gen, type =2)

# Ranking genotype 
plot(gge_model_gen, type=8)

```

```{r genotype-biplots-code-hidden, warning=FALSE, message=FALSE, include=FALSE}

#mean performance vs stability 
mean_vs_stability <- plot(gge_model_gen, type =2, #<<
                          
                          plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14)) +
  ylim(-2000,1500)


# Ranking genotype 
rank_genotype <- plot(gge_model_gen, type=8, #<<
                      
                      plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14)) +
  ylim(-2000,2000)

```
]
.panel[.panel-name[Mean vs Stability]

```{r mean-perf-vs-stability-plot, message=FALSE, warning=FALSE}
mean_vs_stability
```
]
.panel[.panel-name[Ranking Genotypes]
```{r type-8-plot, message=FALSE, warning=FALSE}
rank_genotype
```
]
]


---
class: middle, center

# Finally, for the third model, let's look at svp = "symmetrical". 


---

.panelset[
.panel[.panel-name[Create Biplot]
```{r svp-symmetrical}
gge_symmetrical <- gge(.data = bean_data,
                 env = Site,
                 gen = Gen_id,
                 resp = yield,
                 centering = "environment",
                 svp = "symmetrical")
```

]
.panel[.panel-name[R Code]
```{r symmetrical-biplots-code-hidden, message=FALSE, include=FALSE}
# which won where 
which_won_where <- plot(gge_symmetrical, type=3, #<<
                        
                        
                        size.text.win = 5, plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14)) +
  ylim(-60,25)

#Comparing 2 genotypes 

compare_two_gen <- plot(gge_symmetrical, type=9, sel_gen1 = "15", sel_gen2 = "69",  #<<
                        
                         size.text.win = 6, plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14)) +
  ylim(-60,30)

```
```{r symmetrical-biplots-code, message=FALSE, results='hide', eval=FALSE}
# which won where 
plot(gge_symmetrical, type=3)

#Comparing 2 genotypes 

compare_two_gen <- plot(gge_symmetrical, type=9, sel_gen1 = "15", sel_gen2 = "69")

```
]
.panel[.panel-name[Which Won Where]
```{r type-3-plot, message=FALSE, warning=FALSE}
which_won_where
```
]
.panel[.panel-name[Compare Genotypes]
```{r type-9-plot, message=FALSE, warning=FALSE}
compare_two_gen
```
]
]

---
class: middle

## Additive Main-Effects Multiplicative Interaction (AMMI)

AMMI deals with average performances and specific performances/GxE interaction. 

* Addive Main-Effects:
  * Genotype Main Effect
  * Environment Main Effect
  
* Multiplicative Interaction
  * Genotype * Environment Interaction Effect 

---

##### Creating an AMMI Model

.panelset[
.panel[.panel-name[AMMI Model]
Use performs_ammi() to create an ammi object. From the ammi object, you can get:

* ANOVA table
* PCA 
* AMMI model scores
* means of genotypes in the environments
```{r ammi-model, results= 'hide'}
ammi <- performs_ammi(.data = bean_data, 
                      env = Site, 
                      gen = Gen_id, 
                      rep = Rep,
                      resp = yield)
```
]
.panel[.panel-name[AMMI ANOVA]
```{r ammi-anova, highlight.output = 1:9}
ammi$yield$ANOVA
```
]
.panel[.panel-name[AMMI Plot]

```{r ammi-plot}
residual_plots(ammi)
```
]
]

---

###### AMMI Biplots

```{r creating-plots-hidden, include=FALSE}
options(ggrepel.max.overlaps = 3) 

#
ammi_biplot1 <- plot_scores(ammi, x.lab = "Yield", axis.expand = 1, size.tex.gen = 3, size.tex.env = 5,
                   plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 14,
  title_font_size = 14))

#
b2 <- plot_scores(ammi, type=2,
                  plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14))

#
b2_polygon <- plot_scores(ammi, type=2, polygon=T, 
                          plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14))

#
b2_other <- plot_scores(ammi,
                  type=2,
                  col.env = "green4",
                  col.segm.env = "orange",
                  highlight = c("1", "2"), 
                  col.highlight = "darkcyan",
                  axis.expand = 1.5,
                  plot_theme = theme_xaringan(css_file =  here("GxE-June2022/xaringan-themer.css"),
  text_font_size = 12,
  title_font_size = 14))


```

.panelset[
.panel[.panel-name[Creating the plots]
```{r creating-plots, results='hide', warning=FALSE, eval=FALSE}
#ab2, basic ammi biplot
plot_scores(ammi)

#b2, Produces an AMMI2 biplot (PC1 x PC2) to make inferences related to the interaction effects.
plot_scores(ammi, type=2)

#b2 polygon
plot_scores(ammi, type=2, polygon=T)


```
]
.panel[.panel-name[ammi_biplot1]
```{r ab2, warning=FALSE}
ammi_biplot1

```
]

.panel[.panel-name[b2]
```{r b2, warning=FALSE}
b2

```
]

.panel[.panel-name[b2_polygon]
```{r b2-polygon, warning=FALSE}
b2_polygon

```
]
]

---
class: middle, center

# Ranks: Yield Output and Stability 

#### Ammi Stability Value (ASV) and Yield Stability Index (YSI) are the metrics I use/come across most frequently. Ranking genotypes using these metrics is useful because it takes into consideration both yield output and stability. 

--

.pull-left[
#### ASV: genotypes with low ASV values are more stable and vice versa. 
]

--

.pull-right[
#### YSI takes into account both stability and yield output making it a better method to identify high yielding and stable genotypes. Lower YSI have both high mean yield and stability traits. 
]

---

#### Use the <a href= "https://tiagoolivoto.github.io/metan/reference/ammi_indexes.html">ammi_indexes()</a> function to calculate many AMMI-based statistics.
.panelset[
.panel[.panel-name[ASV]
```{r ASV, results= 'hide'}

ammi_index <- ammi_indexes(ammi, level = 0.95)
```
]
.panel[.panel-name[ASV Results]

```{r ASV-results}

ammi_index$yield$ASV

```
]
.panel[.panel-name[ASVR Results]

```{r ASVR-results}
ammi_index$yield$ASV_R

```
]

.panel[.panel-name[YSI]

### YSI = RASV + RY

]
]


---
class: middle

# Follow Up Resources 

#### The link to this presentation will be emailed to you.

#### Check out my <a href = "https://rforplantbreeders.netlify.app/"> blog </a> for more detailed written tutorials for plant breeders who want to learn how to use R. 

#### Current Posts:

*  <a href = "https://rforplantbreeders.netlify.app/posts/introduction-to-r-for-plant-breeding/"> Introduction to R </a>
*  <a href = "https://rforplantbreeders.netlify.app/posts/importing-data-into-r/"> Importing in R </a>

---
class: middle, center

# Any Questions?
### Thank you for attending this workshop. You can ask any questions you may have now. 

